Galactic Chemical Evolution Models Favor an Extended Type Ia Supernova Delay-Time Distribution

### 2024-03-26 src/tex/ms.tex
[showyourwork] overleaf sync

% Define document class
% \documentclass[modern,linenumbers]{aastex631}
\documentclass[twocolumn,twocolappendix,linenumbers,trackchanges]{aastex631}

@@

@@

The uncertainties in the SN Ia DTD propagate into GCE models. In principle, the observed chemical abundance patterns should therefore contain information about the SN Ia DTD, and by extension their progenitors. A striking feature of the [O/Fe] distribution in the Milky Way is the distinct separation into two components, the high- and low-$\alpha$ sequences [references]. [Introduce other MW disk features].

A few studies have compared different DTDs in one-zone chemical evolution models, but their comparisons to abundance data are limited to the local solar neighborhood \citep[e.g.,][]{Andrews2017-ChemicalEvolution,Palicio2023-AnalyticDTD}. 
[Talk about \citet{Matteucci2009-DTDModels} multi-zone models].
\citet{Poulhazan2018-PrecisionPollution} compared six DTDs in a cosmological smoothed-particle hydrodynamics simulation. They found that models with a prompt SN Ia component produced the highest abundance ratios and the narrowest \aFe distributions at late times, although their simulation was not tuned to reproduce the parameters of the Milky Way, and they did not compare to observed chemical abundances. 
The current era of large spectroscopic surveys such as the Apache Point Observatory Galactic Evolution Experiment \citep[APOGEE;][]{Majewski2017-APOGEE} and the forthcoming Milky Way Mapper \citep{Kollmeier2017-SDSS-V} has made abundances across the Milky Way disk available for comparison to more sophisticated GCE models.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{figures/ofe_bimodality_summary.pdf}
    \caption{The distributions of [O/Fe] along two different slices of [Fe/H]. 
    % $-0.6\leq$[Fe/H]$<-0.4$ (blue dashed) and $-0.4\leq$[Fe/H]$<-0.2$ (red solid). 
    All panels contain stars within the Galactic region defined by $7\leq R_{\rm gal}<9$ kpc and $0\leq|z|<2$ kpc. The $|z|$-distributions of the model stellar populations are re-sampled to match the vertical distribution of the APOGEE sample, and \num{20000} stellar populations are drawn from the adjusted distribution in each panel. The maximum of each distribution is normalized to 1 and the vertical scale is consistent across all panels.
    \textit{Top row:} results from five multi-zone models which assume the late-bust SFH but different DTDs. \textit{Bottom row}: the first four panels compare the four SFHs, all assuming an exponential DTD with $\tau=1.5$ Gyr. The bottom-right panel (highlighted) plots data from APOGEE DR17 for reference.}
    \label{fig:ofe-bimodality}
    \script{ofe_bimodality_summary.py}
\end{figure*}

[Intro, connect to previous paragraph]. The relative prominence of the two peaks varies with $R_{\rm gal}$ and $|z|$, as shown by the right-hand panels of Figures \ref{fig:ofe-df-sfh} and \ref{fig:ofe-df-dtd}. A crucial feature of this bimodality, which is not apparent in the previous figures, is the presence of both sequences at the same [Fe/H]. Figure \ref{fig:ofe-bimodality} compares the [O/Fe] distributions in the solar annulus ($7\leq R_{\rm gal}<9$ kpc and $0\leq|z|<2$ kpc) in two bins of [Fe/H] for select model outputs and APOGEE data. The APOGEE distributions in the bottom-right panel (j) show that the high-$\alpha$ mode is more prominent at lower [Fe/H] and vice versa, but the distributions in both bins are clearly bimodal. Moreover, the ``trough'' at ${\rm [O/Fe]}\approx0.2$ is consistent in each bin.

[Reference \citet{Vincenzo2021-AlphaDistribution} for how the bimodality is made stronger by the selection function, but it's not just due to selection effects. Describe the re-sampling we do to correct for this.]
% \citet{Vincenzo2021-AlphaDistribution} demonstrate that the APOGEE selection function makes the observed $\alpha$ bimodality {\it weaker} in that the high-$\alpha$ sequence is undersampled relative to low-$\alpha$ stars.
For most of our models, the low-$\alpha$ peak is the more prominent of the two, so by performing this resampling we enhance strength of the bimodality. However, for some models the high-$\alpha$ peak is the strongest, and the re-sampling process de-emphasizes the low-$\alpha$ peak below the prominence threshold (e.g., panel f of Figure \ref{fig:ofe-bimodality}). [Explain why this is ok]

The top row of panels (a--e) in Figure \ref{fig:ofe-bimodality} shows the [O/Fe] bimodality (or lack thereof) across five different DTDs, all of which assume the late-burst SFH. To quantify the bimodality, we use the peak-finding algorithm {\tt scipy.signal.find\_peaks} \citep{2020SciPy-NMeth}. We calculate the prominence, or vertical distance between a peak and its highest neighboring trough, and adopt an arbitrary threshold of 0.1. Five of the eight DTDs (four of the five shown, plus the exponential form with $\tau=3$ Gyr) exceed this prominence threshold, while both power-law DTDs and the plateau ($W=0.3$ Gyr) model do not. Panel (b) shows that the power-law ($\alpha=-1.1$) DTD produces a marginal high-$\alpha$ peak, although it doesn't meet the prominence threshold because it is too close to the low-$\alpha$ peak.

In general, DTDs with fewer prompt SNe Ia produce a high-$\alpha$ peak which is more prominent and at a higher [O/Fe], as was the case with the [O/Fe] distributions in Section \ref{sec:ofe-df}. One major discrepancy is that the APOGEE data show bimodality in both bins of [Fe/H], while the models run with the late-burst SFH only show a clear high-$\alpha$ peak at the low-metallicity ($-0.6<{\rm [Fe/H]}<-0.4$) bin. [Talk about why / if this matters]

Panels (f)--(i) in the bottom row of Figure \ref{fig:ofe-bimodality} illustrate the effect of the SFH on the [O/Fe] bimodality with the exponential ($\tau=1.5$ Gyr) DTD. The inside-out SFH does not produce a bimodal distribution for any of our DTDs. On the other hand, the early-burst SFH {\it always} produces a bimodal distribution in the $-0.4\leq{\rm [Fe/H]}<-0.2$ bin regardless of the assumed DTD, although the lower-metallicity bin does not (the lower peak falls short of our prominence threshold of 0.1).

For models with the late-burst and two-infall SFHs, bimodality is variable depending on the DTD: those with longer median delay times (e.g., exponential, plateau, or triple-system) generally produce a bimodal distribution in the $-0.6\leq{\rm [Fe/H]}<-0.4$ bin, while the power-law DTDs do not. None of our SFH+DTD combinations can produce an [O/Fe] distribution which is bimodal in {\it both} bins of [Fe/H] shown in Figure \ref{fig:ofe-bimodality}, as observed in APOGEE (panel j).

@@

@@

[Is the MW DTD the same as elsewhere?]
\citet{Walcher2016-SelfSimilarity} argue that the similarity of the age--[$\alpha$/Fe] relation between solar neighborhood stars and nearby elliptical galaxies is evidence for a universal DTD; between a bimodal, narrow Gaussian, and power-law DTD, they find the latter best explains their data.

@@

@@

One curious chemical feature of the Milky Way disk is the existence of two populations separated by their \aFe abundance ratios at similar metallicity \citep[e.g.,][]{Bensby2014-solarNeighborhoodAbundances}.

@@

\begin{acknowledgments}
    We thank Prof.\ David H.\ Weinberg and attendees of OSU's Galaxy Hour for many useful discussions over the course of this project.
    [Personal acknowledgements]

@@

    [Carnegie land acknowledgement]
    % Software acknowledgements if not a AAS journal
    % This work made use of Astropy:\footnote{http://www.astropy.org} a community-developed core Python package and an ecosystem of tools and resources for astronomy \citep{astropy:2013, astropy:2018, astropy:2022}. 
\end{acknowledgments}

>>>>

% Define document class
% \documentclass[modern,linenumbers]{aastex631}
% \documentclass[twocolumn,twocolappendix,linenumbers,trackchanges]{aastex631}
\documentclass[twocolumn,twocolappendix,linenumbers]{aastex631}

@@

\correspondingauthor{Liam O. Dubay}
\email{dubay.11@osu.edu}

@@

The uncertainties in the SN Ia DTD propagate into GCE models. In principle, the observed chemical abundance patterns should therefore contain information about the DTD, and by extension the progenitors of SNe Ia. The metallicity distribution function (MDF) and distribution of [O/Fe] record the history of SN Ia enrichment as a function of stellar age and location in the Galaxy. A striking feature of the \aFe distribution in the Milky Way disk is the distinct separation into two components, the high- and low-$\alpha$ sequences, at similar metallicity \citep[e.g.,][]{Bensby2014-solarNeighborhoodAbundances}. Since the \aFe abundance reflects the ratio of CC SN to SN Ia enrichment, the DTD should have an effect on the \aFe bimodality.
% [Introduce other MW disk features?]

A few studies have compared different DTDs in one-zone chemical evolution models, but their comparisons to abundance data are limited to the local solar neighborhood \citep[e.g.,][]{Andrews2017-ChemicalEvolution,Palicio2023-AnalyticDTD}. =
\citet{Poulhazan2018-PrecisionPollution} compared six DTDs in a cosmological smoothed-particle hydrodynamics simulation. They found that models with a prompt SN Ia component produced the highest abundance ratios and the narrowest \aFe distributions at late times, although their simulation was not tuned to reproduce the parameters of the Milky Way.%, and they did not compare to observed chemical abundances. 
\citet{Matteucci2009-DTDModels} compared multiple DTDs in a multi-zone GCE model with a two-infall SFH, but were limited by the available data for the solar neighborhood.
The current era of large spectroscopic surveys such as the Apache Point Observatory Galactic Evolution Experiment \citep[APOGEE;][]{Majewski2017-APOGEE} and the forthcoming Milky Way Mapper \citep{Kollmeier2017-SDSS-V} has made abundances across the Milky Way disk available for comparison to more sophisticated GCE models.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{figures/ofe_bimodality_summary.pdf}
    \caption{The distributions of [O/Fe] along two different slices of [Fe/H]: $-0.6\leq$[Fe/H]$<-0.4$ (blue dashed) and $-0.4\leq$[Fe/H]$<-0.2$ (red solid). 
    All panels contain stars within the Galactic region defined by $7\leq R_{\rm gal}<9$ kpc and $0\leq|z|<2$ kpc. The $|z|$-distributions of the model stellar populations are re-sampled to match the vertical distribution of the APOGEE sample, and \num{20000} stellar populations are drawn from the adjusted distribution in each panel. The maximum of each distribution is normalized to 1 and the vertical scale is consistent across all panels.
    \textit{Top row:} results from five multi-zone models which assume the late-bust SFH but different DTDs. \textit{Bottom row}: the first four panels compare the four SFHs, all assuming an exponential DTD with $\tau=1.5$ Gyr. The bottom-right panel (highlighted) plots data from APOGEE DR17 for reference.}
    \label{fig:ofe-bimodality}
    \script{ofe_bimodality_summary.py}
\end{figure*}

% [Intro, connect to previous section]. 
The [O/Fe] distributions from APOGEE in Section \ref{sec:ofe-df}
% (right-hand panels of Figures \ref{fig:ofe-df-sfh} and \ref{fig:ofe-df-dtd}) 
show two distinct peaks whose relative prominence varies with $R_{\rm gal}$ and $|z|$. A crucial feature of this bimodality, which is not apparent in the analysis of the previous section, is the presence of both sequences {\it at the same [Fe/H]}. \citet{Vincenzo2021-AlphaDistribution} demonstrate that, although the APOGEE sample over-represents stars with $|z|\ge1$ which are more likely to be high-$\alpha$, the $\alpha$-bimodality is a real feature of the solar neighborhood population. A successful model for the evolution of the Milky Way must be able to reproduce this bimodality.

Figure \ref{fig:ofe-bimodality} compares the [O/Fe] distributions in the solar annulus ($7\leq R_{\rm gal}<9$ kpc and $0\leq|z|<2$ kpc) in two bins of [Fe/H] ($-0.6<{\rm [Fe/H]}<-0.4$ and $-0.4<{\rm [Fe/H]}<-0.2$) for select model outputs and APOGEE data. The APOGEE distributions in the bottom-right panel (j) show that the high-$\alpha$ mode is more prominent at lower [Fe/H] and vice versa, but the distributions in both bins are clearly bimodal. Moreover, the ``trough'' at ${\rm [O/Fe]}\approx0.2$ is consistent in each bin.

To quantify the strength of the $\alpha$-bimodality, we use the peak-finding algorithm {\tt scipy.signal.find\_peaks} \citep{2020SciPy-NMeth}. For each peak, we calculate the prominence, or the vertical distance between a peak and its highest neighboring trough. We consider a distribution bimodal if both peaks exceed an arbitrary threshold of 0.1. The APOGEE distributions exceed this threshold in both [Fe/H] bins.

The top row of panels (a--e) in Figure \ref{fig:ofe-bimodality} shows the [O/Fe] bimodality (or lack thereof) across five different DTDs, all of which assume the late-burst SFH. To better approximate the APOGEE selection function, we re-sample our model stellar populations so the $|z|$ distribution closely matches that of APOGEE in the solar neighborhood. 
% For most of our models, the low-$\alpha$ peak is the more prominent of the two, so by performing this resampling we enhance strength of the bimodality. However, for some models the high-$\alpha$ peak is the strongest, and the re-sampling process de-emphasizes the low-$\alpha$ peak below the prominence threshold (e.g., panel f of Figure \ref{fig:ofe-bimodality}). [Explain why this is ok]
Five of the eight DTDs (four of the five shown, plus the exponential form with $\tau=3$ Gyr) exceed our prominence threshold in the low-[Fe/H] bin, while both power-law DTDs and the plateau ($W=0.3$ Gyr) model do not. Panel (b) shows that the power-law ($\alpha=-1.1$) DTD produces a marginal high-$\alpha$ peak, although it doesn't meet the prominence threshold because it is too close to the low-$\alpha$ peak. In general, DTDs with fewer prompt SNe Ia produce a high-$\alpha$ peak which is more prominent and at a higher [O/Fe], as was the case with the [O/Fe] distributions in Section \ref{sec:ofe-df}.

Panels (f)--(i) in the bottom row of Figure \ref{fig:ofe-bimodality} illustrate the effect of the SFH on the [O/Fe] bimodality. The inside-out SFH does not produce a bimodal distribution for any of our DTDs. On the other hand, the early-burst SFH {\it always} produces a bimodal distribution in the high-[Fe/H] bin regardless of the assumed DTD, but not in the low-[Fe/H] bin (the small low-$\alpha$ peak falls below our prominence threshold). For models with the late-burst and two-infall SFHs, the bimodality is variable depending on the DTD: those with longer median delay times (e.g., exponential, plateau, or triple-system) generally produce a bimodal distribution in the low-[Fe/H] bin, while the power-law DTDs do not.

One major discrepancy in all of our models the breadth of the $\alpha$-bimodality across [Fe/H] space. 
% In APOGEE, both bins in [Fe/H] show a clearly bimodal distribution of [O/Fe]. 
Even our most successful models can produce a bimodal [O/Fe] distribution in only one bin: the high-[Fe/H] bin for the early-burst SFH, and the low-[Fe/H] bin for the late-burst and two-infall SFHs. In APOGEE, the two sequences are co-extant between ${\rm [Fe/H]}\approx-0.6$, below which the high-$\alpha$ sequence dominates, and ${\rm [Fe/H]}\approx+0.2$, at which point they join. The failure of our models to fully reproduce the bimodality across the whole range of [Fe/H] indicates that our picture of the formation of the $\alpha$-bimodality is still incomplete.

@@

[Age at the knee, given our age uncertainties, is not a great diagnostic]

@@

% [Is the MW DTD the same as elsewhere?]
Our finding that the $t^{-1}$ power-law DTD poorly reproduces observed abundances could signal that the Milky Way and other galaxies obey different DTDs. This would not be too far beyond the findings of \citet{Maoz2017-CosmicDTD} that field galaxies and galaxy clusters have a different DTD slope. On the other hand, \citet{Walcher2016-SelfSimilarity} argue that the similarity of the age--\aFe relation between solar neighborhood stars and nearby elliptical galaxies is evidence for a universal DTD. A physical mechanism would be needed to produce different DTDs in different environments, such as a metallicity dependence in the fraction of close binaries \citep[e.g.,][]{Moe2019-CloseBinaryFraction}.

[Discuss yield effects]

@@

[More discussion of \citet{Matteucci2009-DTDModels}, which is multi-zone but only compares to the solar neighborhood, and no radial migration.]

@@

% One curious chemical feature of the Milky Way disk is the existence of two populations separated by their \aFe abundance ratios at similar metallicity \citep[e.g.,][]{Bensby2014-solarNeighborhoodAbundances}.

@@

% \begin{acknowledgments}
\section*{Acknowledgements}

We thank Prof.\ David H.\ Weinberg and attendees of OSU's Galaxy Hour for many useful discussions over the course of this project.
[Personal acknowledgements]

@@

[Carnegie land acknowledgement]
% Software acknowledgements if not a AAS journal
% This work made use of Astropy:\footnote{http://www.astropy.org} a community-developed core Python package and an ecosystem of tools and resources for astronomy \citep{astropy:2013, astropy:2018, astropy:2022}. 
% \end{acknowledgments}
